---
title: "Current COVID-19 and Vaccination Status in Chile"
description: |
  As on 31 August 2021, 1,638,675 confirmed cases and 36,937 deaths have been recorded, however, the good news is over 70% of the population in Chile is now fully vaccinated. Data from Our World in Data, retrieved on 1 Sep 2021.
categories:
  - Chile
  - Cases and Vaccination
preview: covid.jpg
author:
  - name: Rachel Xinrui WANG
    affiliation: MBAt - Monash Univeristy
date: 09-01-2021
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r library, include=FALSE}
library(tidyverse)
library(formattable)
library(DT)
library(gt)
```

```{r data, include=FALSE}
chile <- read_csv("chile.csv") %>% 
  select(-...1)
```

```{r tabledf, include=FALSE}
monthly <- chile %>%
  separate(col = date,
           into = c("year", "month", "day"),
           sep = "-",
           remove = TRUE) %>% 
  filter(year == 2021) %>% 
  select(-c(year, day)) %>% 
  group_by(month) %>% 
  summarise(new_cases = sum(new_cases, na.rm = TRUE),
            total_cases = max(total_cases, na.rm = TRUE),
            new_deaths = sum(new_deaths, na.rm = TRUE),
            total_deaths = max(total_deaths, na.rm = TRUE),
            new_tests = sum(new_tests, na.rm = TRUE),
            total_tests = max(total_tests, na.rm = TRUE),
            avg_positive = mean(positive_rate, na.rm = TRUE)*100,
            avg_tests = mean(tests_per_case, na.rm = TRUE),
            new_vac = sum(new_vaccinations, na.rm = TRUE),
            total_vac = max(total_vaccinations, na.rm = TRUE),
            pp_vac = max(people_vaccinated, na.rm = TRUE),
            vac_rate = max(people_vaccinated_per_hundred, na.rm = TRUE),
            pp_fullvac = max(people_fully_vaccinated, na.rm = TRUE),
            fullvac_rate = max(people_fully_vaccinated_per_hundred, na.rm = TRUE)) %>% 
  ungroup()

monthly$month[which(monthly$month == "01")] = "Jan"
monthly$month[which(monthly$month == "02")] = "Feb"
monthly$month[which(monthly$month == "03")] = "Mar"
monthly$month[which(monthly$month == "04")] = "Apr"
monthly$month[which(monthly$month == "05")] = "May"
monthly$month[which(monthly$month == "06")] = "Jun"
monthly$month[which(monthly$month == "07")] = "Jul"
monthly$month[which(monthly$month == "08")] = "Aug"


tabledf <-monthly %>% 
  select(month,
         new_cases,
         total_cases,
         new_deaths,
         total_deaths,
         pp_vac,
         pp_fullvac) %>% 
  rename("Month" = month,
         "New Cases" = new_cases,
         "Total Confirmed Cases" = total_cases,
         "New Deaths" = new_deaths,
         "Total Deaths" = total_deaths,
         "People Vaccinated" = pp_vac,
         "People Fully Vaccinated" = pp_fullvac)

```

After fighting against COVID-19 for almost a year, vaccine became available in January 2021, and made significant contributions in protecting people from the virus. The table below demonstrates a monthly summary regarding average number of tests performed per confirmed case, average percentage of positive test results, proportion of the population who are vaccinated and fully vaccinated in Chile since the vaccine was introduced.

```{r table1}
tabledf1 <- monthly %>% 
  select(month,
         avg_tests,
         avg_positive,
         vac_rate,
         fullvac_rate) %>% 
  gt() %>% 
  tab_header(title = "Average Positive Rate by Tests Dropped to 1.52% with Over 70% of the Population Fully Vaccinated",
             subtitle = "Most Recent Updates Recorded in the End of August 2021") %>% 
  cols_align(align = "center",
             columns = 1:5) %>% 
  fmt_number(columns = c(avg_tests,
                         avg_positive),
             decimals = 2) %>% 
  tab_source_note(md("Source: 2021 *Our World in Data*")) %>% 
  cols_label(month = html("Month"),
             avg_tests = html("Average Tests per Case"),
             avg_positive = html("Average Positive Rate<br>(%)"),
             vac_rate = html("Vaccinated<br>(%)"),
             fullvac_rate = html("Fully Vaccinated<br>(%)"))
tabledf1
```

The average positive rate dropped from 7.52% in January when only 0.3% of the population is vaccinated, to 1.52% in August where nearly 75% of the population vaccinated, and over 70% are fully vaccinated. Average number of tests conducted to detect a positive case also increased from 13.34 to 68.26 per month, suggests that by the end of August in 2021, only 1 person may test positive in 69 tests. The statistics above shows a very encouraging trend of the COVID-19 situation in Chile, **the spread of COVID-19 virus is getting under control while majority of the population is now fully vaccinated.**

A clearer picture can be observed from the table below regarding the vaccination progress as well as the cumulative number of confirmed cases, tests conducted and deaths since January 2021. By the end of August 2021, over 14 million people in Chile had their first vaccine and over 13 million are fully vaccinated, the increasing trend of confirmed cases slowed down. An obvious decreasing trend can be observed from April, when the vaccination rate reached nearly 50%, the number of new confirmed cases per month dropped from over 200,000 to only one tenth of this figure in August at slightly over 20,000. The number of deaths each month also witnessed a relatively steady decrease but not as much as in the number of new cases, which indicates that **although vaccine effectively protected people from the virus, treatments for those who already infected by COVID-19 still needs further improvements to reduce the number of deaths.** 

As on 31 August 2021, 1,638,675 confirmed cases have been recorded in total, with 36,937 deaths, these figures are still concerning even though the overall COVID-19 situation in Chile is getting better with majority of the population fully vaccinated. The number of new deaths and daily confirmed cases are expected to continue decreasing while more people are getting vaccinated and more research conducted in COVID-19 treatments.

```{r table2}
formattable(tabledf, 
            align = c("l", "c", "c", "c", "c", "c", "c"),
            list(`Month` = formatter("span",
                                    style = ~style(color = "grey",
                                                   font.weight = "bold")),
                 `New Cases` = color_tile("transparent", "tomato"),
                 `Total Confirmed Cases` = color_tile("transparent", "steelblue3"),
                 `New Deaths` = color_tile("transparent", "lightcoral"),
                 `Total Deaths` = color_tile("transparent", "steelblue3"),
                 `People Vaccinated` = color_tile("transparent", "yellowgreen"),
                 `People Fully Vaccinated` = color_tile("transparent", "limegreen")))
```

Historical data of daily and cumulative confirmed cases, deaths, tests performed, vaccination administrated and number of people vaccinated from February 2020 to August 2021 is also available below for anyone who is interested.

```{r, include=FALSE}
tabledf2 <- chile %>% 
  select(date,
         new_cases,
         total_cases,
         new_deaths,
         total_deaths,
         new_tests,
         total_tests,
         new_vaccinations,
         total_vaccinations,
         people_vaccinated,
         people_vaccinated_per_hundred,
         people_fully_vaccinated,
         people_fully_vaccinated_per_hundred)
```

```{r table3}
datatable(tabledf2,
          colnames = c("Date", 
                       "New Cases", 
                       "Total Cases", 
                       "New Deaths",
                       "Total Deaths",
                       "New Tests", 
                       "Total Tests", 
                       "New Vaccinations", 
                       "Total Vaccinations",
                       "People Vaccinated", 
                       "Vaccinated Rate", 
                       "People Fully Vaccinated",
                       "Fully Vaccinated Rate"),
          options = list(initComplete = JS(
          "function(settings, json) {",
          "$(this.api().table().header()).css({'background-color': '#404b5d', 'color': '#fff'});",
          "}"),
          pageLength = 7,
          scrollX = TRUE,
          columnDefs = list(list(className = "dt-center", targets = 2:13),
                            list(className = "dt-left", targets = 1)))
          )
```

